---
title: "Using nimbleSCR to fit a point process SCR model to wolverine (Gulo gulo) non-invasive genetic sampling data"
author: "Pierre Dupont"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{Wolverine SCR_pp Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 8, fig.height = 10) 
```

```{r echo = FALSE}
## So it works for everyone
if(Sys.info()['user'] == 'dturek') {
  baseDir <- '~/github/nimble/nimbleSCR/'                   ## Daniel
} else if(Sys.info()['user'] == 'pidu') {
  baseDir <- 'C:/Users/pidu/PACKAGES/nimbleSCR/'            ## Pierre
} else if(Sys.info()['user'] == 'cymi') {
  baseDir <- 'C:/Personal_Cloud/OneDrive/Work/nimbleSCR/'   ## Cyril
} else if(Sys.info()['user'] == 'arichbi') {
  baseDir <- 'C:/PROJECTS/nimbleSCR/'                       ## Richard
} else baseDir <- NULL

if(is.null(baseDir)) {
    load(system.file("extdata", "WolverinePointProcess_data.RData", package = "nimbleSCR"))
} else {
    load(file.path(baseDir,"nimbleSCR/inst/extdata/WolverinePointProcess_data.RData"))
}
```


This vignette demonstrates the wolverine point-process SCR model, as in
"A flexible and efficient Bayesian implementation of point process models for 
spatial capture-recapture data" (Wei *et al*, *submitted*) using NIMBLE [@de2017programming;@nimbleSoftware2020] and the nimbleSCR package [@nimbleSCR].


### Load Libraries 

```{r , warning = FALSE, message = FALSE}
library(nimble)
library(nimbleSCR)
library(basicMCMCplots)
```


### Define Model Structure

Here, we define the `nimble` model structure.
```{r }
modelCode <- nimbleCode({  
  ##------ SPATIAL PROCESS 
  ## Intercept and slope for the log-linear model for habitat selection intensity
  habCoeffInt ~ dnorm(0, sd = 10)
  habCoeffSlope ~ dnorm(0, sd = 10)
  
  ## Habitat intensity for each habitat window
  habIntensity[1:numHabWindows] <- exp(habCoeffInt + habCoeffSlope * habCovs[1:numHabWindows])
  sumHabIntensity <- sum(habIntensity[1:numHabWindows])
  logHabIntensity[1:numHabWindows] <- log(habIntensity[1:numHabWindows])
  logSumHabIntensity <- log(sumHabIntensity)
  
  ## Activity centres of the observed individuals: a bernoulli point process
  for(i in 1:numIdDetected){
    sxy[i,1:2] ~ dbernppAC(
      lowerCoords = habLoCoords[1:numHabWindows, 1:2],
      upperCoords = habUpCoords[1:numHabWindows, 1:2],
      logIntensities = logHabIntensity[1:numHabWindows],
      logSumIntensity = logSumHabIntensity,
      habitatGrid = habitatGrid[1:y.max,1:x.max],
      numGridRows = y.max,
      numGridCols = x.max)
  }#i
  
  

  ##----- DEMOGRAPHIC PROCESS
  ## Number of individuals in the population
  N ~ dpois(sumHabIntensity)
  
  ## Number of detected individuals
  nDetectedIndiv ~ dbin(probDetection, N)
  
  

  ##----- DETECTION PROCESS 
  ## Scale for the multivariate normal detection function
  sigma ~ dunif(0,2)
  ## Intercept and slope parameters for the log-linear model for detection intensity
  for(c in 1:numCounties){
    detCoeffInt[c] ~ dnorm(0, sd = 10)
  }#c
  for(cc in 1:numDetCovs){
    detCoeffSlope[cc] ~ dnorm(0, sd = 10)
  }#c
  
  ## Baseline detection intensity for each detection window
  for(j in 1:numDetWindows){
    detIntensity[j] <- exp( detCoeffInt[detCounties[j]] +
                              detCoeffSlope[1] * detCovs[j,1] +
                              detCoeffSlope[2] * detCovs[j,2] +
                              detCoeffSlope[3] * detCovs[j,3])
  }#j

  ## Detections of the observed individuals conditional on their activity centers
  for(i in 1:numIdDetected) {
    y[i,1:(maxDetections+1),1:3] ~ dpoisppDetection_normal(
      lowerCoords = detLoCoords[1:numDetWindows,1:2],
      upperCoords = detUpCoords[1:numDetWindows,1:2],
      s = sxy[i,1:2],
      sd = sigma,
      baseIntensities = detIntensity[1:numDetWindows],
      numMaxPoints = maxDetections,
      numWindows = numDetWindows,
      indicator = 1)
  }#i

  ## The probability that an individual in the population is detected at least once
  ## i.e. one minus the void probability over all detection windows
  probDetection <- 1 - marginalVoidProbNumIntegration(
    quadNodes = quadNodes[1:maxNumNodes,1:2,1:numHabWindows],
    quadWeights = quadWeights[1:numHabWindows],
    numNodes = numNodes[1:numHabWindows],
    lowerCoords = detLoCoords[1:numDetWindows,1:2],
    upperCoords = detUpCoords[1:numDetWindows,1:2],
    sd = sigma,
    baseIntensities = detIntensity[1:numDetWindows],
    habIntensities = habIntensity[1:numHabWindows],
    sumHabIntensity = sumHabIntensity,
    numObsWindows = numDetWindows,
    numHabWindows = numHabWindows
  )
  logDetProb <- log(probDetection)
  normData ~ dnormalizer(logNormConstant = logDetProb)

})
```


### Load Wolverine Data

We load the wolverine example data available from [needs to be updated when the paper is published]
The data file can be downloaded at [needs to be updated when the paper is published].
```{r eval = FALSE}
load("WolverinePointProcess_Data.RData")
```


## Create NIMBLE Model

Now, we can create the `nimble` model object, using the model structure
defined in `modelCode`, and the constants, data, and initial values.
```{r eval = FALSE}
Rmodel <- nimbleModel(modelCode, constants, data, inits)
```


#### Configure and Build MCMC

We configure an MCMC algorithm to the `Rmodel` model object.

We assign MCMC monitors to the different parameters we want to tracks (e.g. $N$, $\sigma$, ...)
```{r message = FALSE, eval = FALSE}
conf <- configureMCMC( Rmodel,
                       monitors = c( "N", "habCoeffInt", "habCoeffSlope",
                                     "detCoeffInt", "detCoeffSlope",
                                     "sigma", "probDetection"),
                       print = FALSE)
Rmcmc <- buildMCMC(conf)
```


#### Compile and Run MCMC

Finally, we compile both the model and MCMC objects and
execute the compiled MCMC for 1000 iterations.
```{r eval = FALSE}
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
MCMC_runtime <- system.time(
  samples <- runMCMC(Cmcmc, niter = 1000)
)
```

```{r eval = FALSE, echo = FALSE}
save(samples,
     MCMC_runtime,
     file = file.path(baseDir,
                      "nimbleSCR/inst/extdata/WolverinePointProcess_samples.RData"))
```

```{r echo = FALSE}
if(is.null(baseDir)) {
    load(system.file("extdata",
                     "WolverinePointProcess_samples.RData",
                     package = "nimbleSCR"))
} else {
    load(file.path(baseDir,
                   "nimbleSCR/inst/extdata/WolverinePointProcess_samples.RData"))
}
```


## Results
Summary of posterior distributions for each parameter:
```{r }
round(samplesSummary(samples), 2)
```

Examine traceplots and posterior distributions:
```{r }
chainsPlot(samples)
```

Obviously, we would need to run multiple longer chains but this might take a while
(runtime for 1000 iterations was approx 12 min).



## References

